Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-02-09 newpaper/paper.tex
stuff

\newcommand{\mit}[1]{\mathit{#1}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\mdo}{\mbf{do}\,}
\newcommand{\ind}{\hspace{1em}}
\newcommand{\bif}{\mbf{if}\,}
\newcommand{\bthen}{\mbf{then}\,}
\newcommand{\belse}{\mbf{else}\,}
\newcommand{\return}{\msf{return}\,}
\newcommand{\lam}{\lambda\,}
\newcommand{\data}{\mbf{data}\,}
\newcommand{\where}{\mbf{where}}
\newcommand{\M}{\msf{M}}
\newcommand{\letrec}{\mbf{letrec}\,}
\newcommand{\of}{\mbf{of}\,}
\newcommand{\go}{\mit{go}}
\newcommand{\add}{\mit{add}}
\newcommand{\letdef}{\mbf{let\,}}

@@

\newcommand{\Int}{\msf{Int}}
\newcommand{\List}{\msf{List}}
\newcommand{\Nil}{\msf{Nil}}
\newcommand{\Cons}{\msf{Cons}}
\newcommand{\Reader}{\msf{Reader}}
\newcommand{\Monad}{\msf{Monad}}
\newcommand{\class}{\msf{class}}
\newcommand{\Functor}{\msf{Functor}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\Statel}{\msf{State}}
\newcommand{\fro}{\leftarrow}
\newcommand{\case}{\mbf{case\,}}

\newcommand{\Lift}{{\Uparrow}}
\newcommand{\spl}{{\sim}}
\newcommand{\ql}{{\langle}}
\newcommand{\qr}{{\rangle}}
\newcommand{\bind}{\mathbin{>\!\!>\mkern-6.7mu=}}

@@

\newcommand{\Nat}{\msf{Nat}}
\newcommand{\Zero}{\msf{Zero}}
\newcommand{\Suc}{\msf{Suc}}

%% \newcommand{\spl}{{\sim}}
%% \newcommand{\qut}[1]{\langle #1\rangle}

%% \newcommand{\mbbc}{\mbb{C}}
%% \newcommand{\mbbo}{\mbb{O}}
%% \newcommand{\ob}{_\mbbo}

%% \newcommand{\U}{\msf{U}}
%% \newcommand{\Con}{\msf{Con}}
%% \newcommand{\Sub}{\msf{Sub}}
%% \newcommand{\Ty}{\msf{Ty}}
%% \newcommand{\Tm}{\msf{Tm}}
%% \newcommand{\Cono}{\msf{Con}_{\mbbo}}
%% \newcommand{\Subo}{\msf{Sub}_{\mbbo}}
%% \newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
%% \newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
%% \newcommand{\hCon}{\wh{\msf{Con}}}
%% \newcommand{\hSub}{\wh{\msf{Sub}}}
%% \newcommand{\hTy}{\wh{\msf{Ty}}}
%% \newcommand{\hTm}{\wh{\msf{Tm}}}

%% \newcommand{\p}{\mathsf{p}}
%% \newcommand{\q}{\mathsf{q}}

%% \newcommand{\refl}{\msf{refl}}
%% \newcommand{\Bool}{\msf{Bool}}
%% \newcommand{\true}{\msf{true}}
%% \newcommand{\false}{\msf{false}}
%% \newcommand{\True}{\msf{True}}
%% \newcommand{\False}{\msf{False}}
%% \newcommand{\List}{\msf{List}}
%% \newcommand{\nil}{\msf{nil}}
%% \newcommand{\cons}{\msf{cons}}

%% \renewcommand{\tt}{\msf{tt}}
%% \newcommand{\fst}{\msf{fst}}
%% \newcommand{\snd}{\msf{snd}}
%% \newcommand{\mylet}{\msf{let}}
%% \newcommand{\emptycon}{\scaleobj{.75}\bullet}
%% \newcommand{\id}{\msf{id}}

%% \newcommand{\Set}{\mathsf{Set}}
%% \newcommand{\Prop}{\mathsf{Prop}}
%% \newcommand{\Rep}{\msf{Rep}}

%% \newcommand{\emb}[1]{\ulcorner#1\urcorner}

%% \newcommand{\Stage}{\msf{Stage}}
%% \newcommand{\hato}{\bm\hat{\mbbo}}
%% \newcommand{\ev}{\mbb{E}}
%% \newcommand{\re}{\mbb{R}}

%% \theoremstyle{remark}
%% \newtheorem{notation}{Notation}

%% \newcommand{\whset}{\wh{\Set}}
%% \newcommand{\rexti}{\re_{\ext_1}^{-1}}
%% \newcommand{\rextizero}{\re_{\ext_0}^{-1}}

%% \newcommand{\rel}{^{\approx}}
%% \newcommand{\yon}{\msf{y}}

@@

Metaprograms can be deterministic, transparent, and can be run efficiently,
using fast interpreters or machine code compilation. In contrast,
general-purpose optimizers are slower to run, less transparent and less
reliable. Also, metaprogramming allows \emph{library authors} to exploit domain
specific optimizations, while it is not realistic for general-purpose optimizers
to know about all domains.

@@

Formally, $\MTy$ is additionally indexed by universes levels, and we have
$\MTy_i : \MTy_{i+1}$. However, this adds a bit of noise, and it is not very
relevant to the current paper, so we shall omit levels. Note that universe
levels have nothing to do with staging in the sense of staged compilation; they
are about sizing for the purpose of logical consistency.

We use Agda-style syntax and implicit arguments. A basic example:
\begin{alignat*}{3}
  &id : \{A : \MTy\} \to A \to A\\
  &id = \lam x.\, x
\end{alignat*}
Here, the type argument is implicit, and it gets inferred when we use the
function. For example, $id\,\True$ is elaborated to $id\,\{\Bool\}\,\True$,
where the braces mark an explicit application for the implicit argument.
Similarly, we can introduce implicit lambdas explicitly:
\begin{alignat*}{3}
  &id = \lam \{A : \MTy\}(x : A).\,x
\end{alignat*}
We write $(a : A) \times B$ for $\Sigma$-types in $\MTy$. For the field
projections, for $t : (a : A) \times B$, we have $\fst\,t : A$ and $\snd\,t :
B[a \mapsto \fst\,t]$, and pairing is written simply as $(t,\,u)$. Additionally,
we use syntactic sugar for field projections: we can write a type as $(field_1 :
A) \times (field_2 : B) \times (field_3 : C)$, and write $t.field_2$ for a named
field projection. The type itself denotes a right-nested $\Sigma$-type. We can
also write $(t,\,u,\,v)$ for a right-nested iterated pairing. In Haskell style,
we write $()$ for the unit type, and also for its inhabitant.

Inductive types can be introduced using a Haskell-like ADT notation, or with a GADT-style one:
\begin{alignat*}{3}
  &                                           &&\hspace{4em}\data\,\Bool_\M &&: \MTy\,\where\\
  & \data\,\Bool_\M : \MTy = \True\,|\,\False &&\hspace{4em}\ind\ind \True &&: \Bool_\M\\
  &                                           &&\hspace{4em}\ind\ind \False &&: \Bool_\M
\end{alignat*}
Note that we added an $_\M$ subscript to the type; when analogous types can be
defined both on the meta and object levels, we will use this subscript to
distinguish the meta-level version.

@@

$\bs{\Ty}$ is the universe of types in the object language. It is itself a
metatype, so so we have $\Ty : \MTy$. It is further split to two sub-universes.

First, $\bs{\VTy} : \MTy$ is the universe of \emph{value types}. $\VTy$ supports
parameterized algebraic data types, with two restrictions:
\begin{itemize}
\item ADTs can be only parameterized over types in $\VTy$.
\item Constructor fields must be in $\VTy$.
\end{itemize}
Since $\VTy$ is a sub-universe of $\Ty$, we have that when $A : \VTy$ then also
$A : \Ty$. Formally, this is specified as an explicit embedding operation but
informally it is nicer to have an implicit subtyping.

Second, $\bs{\CTy} : \MTy$ is the universe of \emph{computation types}. This is
also a sub-universe of $\Ty$ with implicit coercions. For now, we only specify
that $\CTy$ contains functions whose domain is a value type:
\[ \blank\to\blank : \VTy \to \Ty \to \CTy \]
For instance, if $\Bool : \VTy$ is defined as an object-level ADT, then $\Bool
\to \Bool : \CTy$, hence also $\Bool \to \Bool : \Ty$. However, $(\Bool \to
\Bool) \to \Bool$ is ill-formed, since the domain is not a value type. An
example for an object-level program, where we already have natural numbers
declared as $\data \Nat : \VTy := \Zero\,|\,\Suc\,\Nat$:
\begin{alignat*}{3}
  &\rlap{$\add : \Nat \to \Nat \to \Nat$}\\
  &\add :=\,&& \mathrlap{\letrec \go := \lam n\,m.\,\case n\,\of}\\
  &         && \ind \Zero   &&\to m;\\
  &         && \ind \Suc\,n &&\to \Suc\,(\go\,n\,m);\\
  &         && \go
\end{alignat*}
Every recursive definition must be introduced with $\letrec$. Note also the
single semicolon after the $\mbf{case}$ branches. The general syntax is $\letrec x :=
t; u$, optionally with a type annotation for $x$ on a separate line.

Also note that object-level definitions use $:=$ instead of $=$ for meta-level
ones. Generally, let-definitions are the biggest source of stage ambiguity in
surface notation. Later on, we will use more implicit staging notation, but we
will always disambiguate the stages of $\mbf{let}$-s in this way. Non-recursive
$\mbf{let}$ is also allowed, and can be used to shadow binders:
\begin{alignat*}{3}
  &f : \Nat \to \Nat\\
  &f := \lam x.\,\letdef x := x + 10; \letdef x := x + 20; x * 10
\end{alignat*}
$\mbf{let}$-definitions can be used to define values of any type, and the type
of the $\mbf{let}$ body can be also anything. Additionally, the right hand sides
of $\case$ branches can also have arbitrary types. So the following is
well-formed:
\begin{alignat*}{3}
  &f : \mathrlap{\Bool \to \Nat \to \Nat}\\
  &f \mathrlap{:= \lam b.\,\case b\,\of}\\
  &\ind \True  &&\to \lam x.\, x + 10;\\
  &\ind \False &&\to \lam x.\, x * 10;
\end{alignat*}
Let us discuss the object language. First, notice that there is no polymorphism
or any kind of type dependency. Although we can define lists as $\data \List\,A
: \VTy := \Nil\,|\,\Cons\,A\,(\List\,A)$, the parameterization is just a
shorthand; all concrete instantiations of the type are distinct. This
monomorphism makes it easy to use different memory layouts for different types.
For example, types which look like products may be unboxed. We could also make a
distinction between boxed and unboxed sum types. We do not explore this in
detail, we just note that monomorphic types make it easy to control memory
layouts, and we believe that this is an important part of performance
optimization.

@@

We say that it should not. We make this formal formal in Section \ref{TODO};
here we only give an intuitive explanation. In short, we choose a call-by-name
semantics for functions, which means that the only way we can compute with a
function is by applying it to all arguments and extracting the resulting value.
Hence, the only way to compute with $f$ is to apply it to \emph{two} arguments,
so $f$ is operationally equivalent to the following definition:
\[ \lam b\,x.\,\case b\,\of \True \to x + 10; \False \to x * 10 \]
In Section \ref{TODO} we show that all object programs can be transformed to a
\emph{saturated} form, where the arity of every function call matches the number
of topmost $\lambda$-binders in the function definition.

Why not just make the object language less liberal, e.g.\ by disallowing
$\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
easier or more obvious? There is a trade-off between making the object language
more restricted, and thus easier to compile, and making \emph{metaprogramming}
more convenient for the object language. We will see that the ability to insert
$\mbf{let}$-s without restriction is very convenient in code generation, and
likewise the ability to have arbitrary object expressions in $\mbf{case}$
bodies. In this paper we choose to go with the most liberal object syntax, at
the cost of needing more downstream processing. The call-by-name nature of
computation types requires a bit of a change of thinking from programmers, but
we believe that it is well worth to have it for the metaprogramming convenience.

@@

Finally, one might compare our object language to call-by-push-value
(CBPV). Indeed, we took inspiration from CBPV, and there are similarities, but
also differences. In both systems there is a value-computation distinction, and
values are call-by-value, and computations are call-by-name. However, our object
language allows variable binding at arbitrary types, while CBPV only supports it
at value types. In CBPV, a let-definition for a function is only possible by
first packing it up in a closure value, which clearly does not work for us.
Also, CBPV makes a judgment-level structural distinction between values and
computations, while we use type universes for that. Generally speaking,
type-based restrictions are easier to work with in dependent type theories than
structural restrictions.

>>>>

\newcommand{\mit}[1]{\mathit{#1}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\mdo}{\mbf{do}\,}
\newcommand{\ind}{\hspace{1em}}
\newcommand{\bif}{\mbf{if}\,}
\newcommand{\bthen}{\mbf{then}\,}
\newcommand{\belse}{\mbf{else}\,}
\newcommand{\return}{\mbf{return}\,}
\newcommand{\lam}{\lambda\,}
\newcommand{\data}{\mbf{data}\,}
\newcommand{\where}{\mbf{where}}
\newcommand{\M}{\msf{M}}
\newcommand{\letrec}{\mbf{letrec}\,}
\newcommand{\of}{\mbf{of}\,}
\newcommand{\go}{\mit{go}}
\newcommand{\add}{\mit{add}}
\newcommand{\letdef}{\mbf{let\,}}
\newcommand{\map}{\mit{map}}
\newcommand{\vas}{\mit{as}}
\newcommand{\vbs}{\mit{bs}}
\newcommand{\vxs}{\mit{xs}}
\newcommand{\vys}{\mit{ys}}
\newcommand{\vma}{\mit{ma}}
\newcommand{\vmb}{\mit{mb}}

@@

\newcommand{\Int}{\msf{Int}}
\newcommand{\List}{\msf{List}\,}
\newcommand{\Nil}{\msf{Nil}}
\newcommand{\Cons}{\msf{Cons}}
\newcommand{\Reader}{\msf{Reader}}
\newcommand{\ReaderT}{\msf{ReaderT}}
\newcommand{\Monad}{\msf{Monad}}
\newcommand{\Applicative}{\msf{Monad}}
\newcommand{\class}{\msf{class}}
\newcommand{\Functor}{\msf{Functor}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\Statel}{\msf{State}}
\newcommand{\fro}{\leftarrow}
\newcommand{\case}{\mbf{case\,}}

\newcommand{\Lift}{{\Uparrow}}
\newcommand{\Up}{{\Uparrow}}
\newcommand{\spl}{{\bs{\sim}}}
\newcommand{\ql}{{\bs{\langle}}}
\newcommand{\qr}{{\bs{\rangle}}}
\newcommand{\bind}{\mathbin{>\!\!>\mkern-6.7mu=}}

@@

\newcommand{\Nat}{\msf{Nat}}
\newcommand{\Zero}{\msf{Zero}}
\newcommand{\Suc}{\msf{Suc}}
\newcommand{\Maybe}{\msf{Maybe}}
\newcommand{\MaybeT}{\msf{MaybeT}}
\newcommand{\Nothing}{\msf{Nothing}}
\newcommand{\Just}{\msf{Just}}

\theoremstyle{remark}
\newtheorem{notation}{Notation}
\newcommand{\id}{\mit{id}}
\newcommand{\mup}{\mit{up}}
\newcommand{\mdown}{\mit{down}}
\newcommand{\tyclass}{\mbf{class}}
\newcommand{\instance}{\mbf{instance}\,}
\newcommand{\Improve}{\msf{Improve}}
\newcommand{\Gen}{\msf{Gen}}
\newcommand{\unGen}{\mit{unGen}}
\renewcommand{\Vec}{\msf{Vec}}
\newcommand{\gen}{\mit{gen}}
\newcommand{\genRec}{\mit{genRec}}
\newcommand{\fmap}{<\!\!\$\!\!>}
\newcommand{\ap}{<\!\!*\!\!>}
\newcommand{\runGen}{\mit{runGen}}
\newcommand{\qt}[1]{\ql#1\qr}
\newcommand{\lift}{\mit{lift}}
\newcommand{\liftGen}{\mit{liftGen}}
\newcommand{\MonadGen}{\msf{MonadGen}}
\newcommand{\RA}{\Rightarrow}
\newcommand{\EitherT}{\msf{EitherT}}
\newcommand{\StateT}{\msf{StateT}}
\newcommand{\Identity}{\msf{Identity}}
\newcommand{\runIdentity}{\mit{runIdentity}}
\newcommand{\newtype}{\mbf{newtype}\,}
\newcommand{\runMaybeT}{\mit{runMaybeT}}
\newcommand{\dlr}{\,\$\,}
\newcommand{\ImproveF}{\msf{ImproveF}}
\newcommand{\ExceptT}{\msf{ExceptT}}

@@

Metaprograms can be deterministic, transparent, and can be run efficiently,
using fast interpreters or machine code compilation. In contrast,
general-purpose optimizers are slower to run, less transparent and less
reliable. Also, metaprogramming allows library authors to exploit
domain-specific optimizations, while it is not realistic for general-purpose
optimizers to know about all domains.

@@

Formally, $\MTy$ is additionally indexed by universes levels, and we have
$\MTy_i : \MTy_{i+1}$. However, this adds a bit of noise, and it is not very
relevant to the current paper, so we shall omit levels. Note that universe
levels have nothing to do with staged compilation; they are about sizing for the
purpose of logical consistency.

We use Agda-like syntax for functions and implicit arguments. A basic example:
\begin{alignat*}{3}
  &\id : \{A : \MTy\} \to A \to A\\
  &\id = \lam x.\, x
\end{alignat*}
Here, the type argument is implicit, and it gets inferred when we use the
function. For example, $id\,\True$ is elaborated to $id\,\{\Bool\}\,\True$,
where the braces mark an explicit application for the implicit argument.
Similarly, we can introduce implicit lambdas explicitly:
\begin{alignat*}{3}
  &id = \lam \{A : \MTy\}(x : A).\,x
\end{alignat*}
We write $(a : A) \times B$ for $\Sigma$-types in $\MTy$. For the field
projections, for $t : (a : A) \times B$, we have $\fst\,t : A$ and $\snd\,t :
B[a \mapsto \fst\,t]$, and pairing is written simply as $(t,\,u)$. Additionally,
we use syntactic sugar for field projections: we can write a type as $(field_1 :
A) \times (field_2 : B) \times (field_3 : C)$, and write $t.field_2$ for a named
field projection. The type itself denotes a right-nested $\Sigma$-type. We can
also write $(t,\,u,\,v)$ for a right-nested iterated pairing. In Haskell style,
we write $()$ for the unit type, and also for its inhabitant.

Inductive types can be introduced using a Haskell-like ADT notation, or with a GADT-style one:
\begin{alignat*}{3}
  &                                           &&\hspace{4em}\data\,\Bool_\M &&: \MTy\,\where\\
  & \data\,\Bool_\M : \MTy = \True\,|\,\False &&\hspace{4em}\ind\ind \True &&: \Bool_\M\\
  &                                           &&\hspace{4em}\ind\ind \False &&: \Bool_\M
\end{alignat*}
Note that we added an $_\M$ subscript to the type; when analogous types can be
defined both on the meta and object levels, we will sometimes use this subscript
to disambiguate the meta-level version. We will also use Haskell-like newtype
notation, such as in $\newtype \msf{Wrap}\,A = \msf{Wrap}\,\{\mit{unWrap} : A
\}$.

Importantly, we can only produce meta-level values by induction on meta-level
data. In general, all construction and elimination rules for type formers in
$\MTy$ stay within $\MTy$.

@@

$\bs{\Ty}$ is the universe of types in the object language. It is itself a
metatype, so so we have $\Ty : \MTy$. Similarly as in the case of $\MTy$, all
construction and elimination rules of the object language stay within $\Ty$.
However, we further split $\Ty$ to two sub-universes.

First, $\bs{\VTy} : \MTy$ is the universe of \emph{value types}. $\VTy$ supports
parameterized algebraic data types, where parameters can have arbitrary types,
but all constructor field types must be in $\VTy$.

Since $\VTy$ is a sub-universe of $\Ty$, we have that when $A : \VTy$ then also
$A : \Ty$. Formally, this is specified as an explicit embedding operation but
informally it is nicer to have an implicit subtyping.

Second, $\bs{\CTy} : \MTy$ is the universe of \emph{computation types}. This is
also a sub-universe of $\Ty$ with implicit coercions. For now, we only specify
that $\CTy$ contains functions whose domains are value types:
\[ \blank\to\blank : \VTy \to \Ty \to \CTy \]
For instance, if $\Bool : \VTy$ is defined as an object-level ADT, then $\Bool
\to \Bool : \CTy$, hence also $\Bool \to \Bool : \Ty$. However, $(\Bool \to
\Bool) \to \Bool$ is ill-formed, since the domain is not a value type. Let us look at an
example for an object-level program, where we already have natural numbers
declared as $\data \Nat := \Zero\,|\,\Suc\,\Nat$:
\begin{alignat*}{3}
  &\hspace{-1em}\rlap{$\add : \Nat \to \Nat \to \Nat$}\\
  &\hspace{-1em}\add :=\,&& \mathrlap{\letrec \go\,n\,m := \case n\,\of}\\
  &\hspace{-1em}         && \ind \Zero   &&\to m;\\
  &\hspace{-1em}         && \ind \Suc\,n &&\to \Suc\,(\go\,n\,m);\\
  &\hspace{-1em}         && \go
\end{alignat*}
Every recursive definition must be introduced with $\letrec$. The general syntax
is $\letrec x : A := t; u$, where the $A$ type annotation can be
omitted. $\mbf{letrec}$ can be only used to define computations, not values
(hence, only functions can be recursive so far).

Object-level definitions use $:=$ as notation, instead of the $=$ that is used
for meta-level ones. Later on, we may use some implicit CFTT notations, but we
will always disambiguate the stages of $\mbf{let}$-s in this way. Non-recursive
$\mbf{let}$ is also allowed, and can be used to shadow binders:
\begin{alignat*}{3}
  &f : \Nat \to \Nat\\
  &f\,x := \letdef x := x + 10; \letdef x := x + 20; x * 10
\end{alignat*}
Values are call-by-value at runtime; they are computed eagerly in function
applications and $\mbf{let}$-s. $\mbf{let}$-definitions can be used to define
inhabitants of any type, and the type of the $\mbf{let}$ body can be also
arbitrary. Additionally, the right hand sides of $\case$ branches can also have
arbitrary types. So the following is well-formed:
\begin{alignat*}{3}
  &\hspace{-1em}f : \mathrlap{\Bool \to \Nat \to \Nat}\\
  &\hspace{-1em}f\,b\,\mathrlap{:= \case b\,\of}\\
  &\hspace{-1em}\ind \True  &&\to \lam x.\, x + 10;\\
  &\hspace{-1em}\ind \False &&\to \lam x.\, x * 10;
\end{alignat*}

What about \textbf{definitional equality} at the object level? This is a
distinct notion from the runtime semantics; object programs are embedded in
CFTT, which is dependently typed, and we need to decide definitional equalities
during type checking. The setup is simple: we have no $\beta$ or $\eta$ rules
for object programs at all, nor any rule for $\mbf{let}$-unfolding. The reasons
are the following. First, since the object language has general recursion, most
$\beta$-rules are only valid in the operational semantics up to some
restrictions, and $\beta$-conversion is not decidable to begin with. Second, in
metaprogramming we care about the size and efficiency of generated code, and
these properties are not stable under $\beta$ and $\eta$.  Hence, the sensible
choice is to do metaprogramming up to strict syntactic equality of object
programs.

Let us discuss the object language. First, notice that there is no polymorphism
or any kind of type dependency. Although we can define lists as $\data \List\,A
: \VTy := \Nil\,|\,\Cons\,A\,(\List\,A)$, the parameterization is just a
shorthand; all concrete instantiations of the type are distinct. This
monomorphism makes it easy to use different memory layouts for different types.
For example, types which look like products may be unboxed. We could also make a
distinction between boxed and unboxed sum types. We do not explore this in
detail, we just note that monomorphic types make it easy to control memory
layouts, and we believe that this is an important part of performance
optimization.

@@

We say that it should not. We make this formal formal in Section \ref{TODO};
here we only give an intuitive explanation. In short, we choose a call-by-name
runtime semantics for functions, which means that the only way we can compute
with a function is by applying it to all arguments and extracting the resulting
value.  Hence, the only way to compute with $f$ is to apply it to \emph{two}
arguments, so $f$ is operationally equivalent to $\lam b\,x.\,\case b\,\of \True
\to x + 10; \False \to x * 10$.  In Section \ref{TODO} we show that all object
programs can be transformed to a \emph{saturated} form, where the arity of every
function call matches the number of topmost $\lambda$-binders in the definition
of the called function. Then, the local functions can be compiled either to
top-level functions by lambda-lifting, or if they are only tail-called, left in
place as join points \cite{TODO}.

Why not just make the object language less liberal, e.g.\ by disallowing
$\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
easier or more obvious? There is a trade-off between making the object language
more restricted, and thus easier to compile, and making \emph{metaprogramming}
more convenient. We will see that the ability to insert $\mbf{let}$-s without
restriction is very convenient in code generation, and likewise the ability to
have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
with the most liberal object syntax, at the cost of needing more downstream
processing. The call-by-name nature of computation types requires a bit of a
change of thinking from programmers, but we believe that it is well worth to
have for the metaprogramming convenience.

@@

Finally, one might compare our object language to call-by-push-value
(CBPV). Indeed, we took inspiration from CBPV, and there are similarities, but
also differences. In both systems there is a value-computation distinction, and
values are call-by-value, and computations are call-by-name. However, our object
language allows variable binding at arbitrary types, while CBPV only supports it
at value types. In CBPV, a let-definition for a function is only possible by
first packing it up as a closure value (or: ``thunk''), which clearly does not
work for us.  Also, CBPV makes a judgment-level structural distinction between
values and computations, while we use type universes for that. Generally
speaking, type-based restrictions are easier to work with in dependent type
theories than structural restrictions.

\subsection{Staging}\label{sec:staging}

With what we have seen so far, there is no interaction between the meta and
object levels. We make such interaction possible with \emph{staging operations}.

\begin{itemize}
\item For $A : \Ty$, we have $\Lift A : \MTy$, pronounced as ``lift $A$''. This is
      the type of metaprograms that produce $A$-typed object programs.
\item For $A : \Ty$ and $t : A$, we have $\ql t \qr : \Lift A$, pronounced ``quote $t$''. This
      is the metaprogram which immediately returns $t$.
\item For $t : \Lift A$, we have $\spl t : A$, pronounced ``splice $t$''. This
  inserts the result of a metaprogram into an object term. \emph{Notation:}
  splicing binds stronger than function application, so $f\,\spl x$ is parsed as
  $f\,(\spl x)$.
\item We have $\ql \spl t \qr \equiv t$ and $\spl \ql t \qr \equiv t$ as definitional equalities.
\end{itemize}

A CFTT program is a mixture of object-level and meta-level top-level definitions
and declarations. \textbf{Staging} means running all metaprograms in splices and
inserting their output into object code, keeping all object-level top entries
and discarding all meta-level ones. Thus, staging takes a CFTT program as input,
and produces output which is purely in the object-level fragment, with no
metatypes and metaprograms remaining. The output is guaranteed to be
well-typed. This staging is formalized in detail in \cite{TODO}. In Section
\ref{TODO} we describe the modifications to ibid.\ that are used in this paper.

Let us look at some basic staging examples. Recall the meta-level identity
function; it can be used at the object-level too, by applying it to quoted
terms:
  \[ \letdef n : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
Here, $\id$ is used at type $\Lift \Nat$. During staging, the expression in the
splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is definitionally
the same as $10 + 10$, which is our staging output here. Boolean
short-circuiting is another basic use-case:
\begin{alignat*}{3}
  &\mit{and} : \Up\Bool \to \Up\Bool \to \Up\Bool\\
  &\mit{and}\,x\,y = \ql\case \spl x\,\of \True \to \spl y; \False \to \False\qr
\end{alignat*}
Since the $y$ expression is inlined under a $\case$ branch at every use site, it
is only computed at runtime when $x$ evaluates to $\True$. In many situations,
staging can be used instead of laziness to implement short-circuiting, and with
generally better runtime performance, avoiding the overhead of thunking. Consider
the $\map$ function now:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\map : \{A\,B : \VTy\} \to (\Up A \to \Up B) \to \Up (\List A) \to \Up(\List B)}\\
  & \hspace{-5em}\mathrlap{\map\,f\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Nil           &&\to \Nil;\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,a\,\vas &&\to \Cons\,\spl(f\,\ql a \qr)\,(\go\,\vas);\\
  & \hspace{-5em}\ind\ind\ind\ind\ind\go\,\spl\vas \qr
\end{alignat*}
For example, this can be used as $\letdef f\,\vas : \List \Nat \to \List \Nat :=
\spl(\map\,(\lam x.\,\ql \spl x + 10 \qr)\,\qt{\vas}) $.  This is staged to a recursive
definition where the mapping function is inlined into the $\Cons$ case as
$\Cons\,a\,\vas \to \Cons\,(a + 10)\,(\go\,\vas)$.  Note that $\map$ has to
abstract over value types, since lists can only contain values, not functions. Also, the
mapping function has type $\Up A \to \Up B$, instead of $\Up (A \to B)$. The
former type is often preferable to the latter in staging; the former is
a metafunction with useful computational content, while the latter is merely a
black box that computes object code. If we have $f : \Up(A \to B)$, and $f$ is
staged to $\ql \lam x.\,t\qr$, then $\spl f u$ is staged to an
undesirable ``administrative'' $\beta$-redex $(\lam x.\,t)\,u$.

\section{Staged Programming in CFTT}\label{sec:abstractions-in-cftt}

\subsection{Binding-Time Improvements}\label{sec:binding-time-improvements}

We saw that $\Up A \to \Up B$ is usually preferable to $\Up(A \to B)$. The two
types are actually equivalent up to the runtime semantics of object programs,
and we can convert back and forth in CFTT:
\begin{alignat*}{3}
  &\mup : \Up (A \to B) \to \Up A \to \Up B && \ind\ind \mdown : (\Up A \to \Up B) \to \Up (A \to B) \\
  &\mup\,f\,a = \ql \spl f\, \spl a\qr   && \ind\ind \mdown\,f = \ql \lam a.\,\spl(f\,\ql a \qr) \qr
\end{alignat*}
We can not show internally, using propositional equality, that these functions are
inverses, since we do not have $\beta\eta$-rules for object functions; but we
will not need that in the rest of the paper.

In the staged compilation and partial evaluation literature, the term
\textbf{binding time improvement} is used to refer to such conversions, where
the ``improved'' version supports more compile-time computation. A general
strategy for generating efficient ``fused'' programs, is to try to work as much
as possible with improved representations, and only convert back to object code
at points where runtime control dependencies are unavoidable.

We use a Haskell-style type class for binding time improvements:
\begin{alignat*}{3}
  &\hspace{-8em}\mathrlap{\tyclass\,\Improve\,(A : \Ty)\,(B : \MTy)\,\where} \\
  &\hspace{-8em}\ind \mup   &&: \Up A \to B\\
  &\hspace{-8em}\ind \mdown &&: B \to \Up A
  &\\
  &\hspace{-8em}\mathrlap{\instance \Improve\,(A \to B)\,(\Up A \to \Up B)\,\where\,\,...}
\end{alignat*}
We will use type classes in an informal way, without precisely specifying how
they work. However, in the Haskell adaptation of this paper we do make crucial
use of type classes. There are some necessary differences between the CFTT and
Haskell versions though, which we will summarize in Section \ref{TODO}. Let us
look at improvement for product types now:
\begin{alignat*}{3}
  &\hspace{-4em}\mathrlap{\instance \Improve\,(A,\,B)\,(\Up A,\,\Up B)\,\where}\\
  &\hspace{-4em}\ind \mup\,x   && = (\ql \fst\,\spl x \qr,\, \ql \snd\,\spl x \qr)\\
  &\hspace{-4em}\ind \mdown\,(a,\,b) && = \ql(\spl a,\, \spl b)\qr
\end{alignat*}
Here we overload Haskell-style product type notation for types and pairing, at
both levels (products are definable as a value type at the object level). There
is a problem with this conversion though: $\mup$ uses $x : \Up(A,\,B)$ twice,
which can increase code size and duplicate runtime computations. For example,
$\mdown\,(\mup\,\ql f\,x \qr)$ is staged to $\ql (\fst\,(f\,x),\,\snd\,(f\,x))
\qr$. It would be safer to first $\mbf{let}$-bind an expression with type
$\Up(A,\,B)$, and then only use projections of the newly bound variable. This is
called \textbf{let-insertion} in staged compilation. But it is impossible to use
let-insertion in $\mup$ because the return type is a product in $\MTy$, and we
cannot introduce object binders in meta-level code.
